Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process

### 2023-12-09 report/appendix_transloc.tex
Reorganised manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,floatsintext]{apa7}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev-Hudilainen3},
  pdflang={en-EN},
  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Modelling writing hesitations}
\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage{colortbl}
\usepackage{threeparttable}
\usepackage[normalem]{ulem}
\usepackage[utf8]{inputenc}
\usepackage{icomma}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process}
\author{Jens Roeser\textsuperscript{1}, Mark Torrance\textsuperscript{1}, Rianne Conijn\textsuperscript{2}, \& Evgeny Chukharev-Hudilainen\textsuperscript{3}}
\date{}

\authornote{

Correspondence concerning this article should be addressed to Jens Roeser, 50 Shakespeare St, Nottingham NG1 4FQ. E-mail: \href{mailto:jens.roeser@ntu.ac.uk}{\nolinkurl{jens.roeser@ntu.ac.uk}}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Psychology, Nottingham Trent University, United Kingdom\\\textsuperscript{2} Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands\\\textsuperscript{3} Department of English, Iowa State University, Iowa}

\abstract{
Classical serial models view the writing process as a chain of pauses and writing burst. In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution. We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production. We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions. Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets. Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations. Our results cannot be explain by the serial but are in line with the cascading view of writing.
}

\begin{document}
\maketitle

Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing (Bock \& Ferreira, 2014; Olive, 2014; Van Galen, 1991; Wheeldon \& Konopka, 2023).

Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech. Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language. Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking (Roeser et al., 2019; Torrance \& Nottbusch, 2012). This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader (e.g. Clark \& Fox Tree, 2002). Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication. Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses (Chenoweth \& Hayes, 2001).

The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models. On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses (Conijn et al., 2019; Flower \& Hayes, 1981; Matsuhashi, 1981; Wengelin, 2006). For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning (Baaijen et al., 2012; Medimorec et al., 2017; Medimorec \& Risko, 2017; Roeser et al., 2019). The same is generally observed for spoken utterances (E.-K. Lee et al., 2013; Meyer, 1996; Wheeldon et al., 2013) with some exceptions (Griffin, 2001). We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted. Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.

Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint (Bock \& Ferreira, 2014; Wheeldon \& Konopka, 2023), the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements (Olive, 2014; Van Galen, 1991). The intervals between adjacent keystrokes during typing vary as a function of various factors. Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured. Processing may be serial. This means that the processing cascade operates on one unit at a time. Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output. Therefore longer IKIs represent periods when the writer is planning what to write next. The subsequent burst of fluent output is then the result this planning (Alves \& Limpo, 2015; Hayes, 2012; Kaufer et al., 1986; Matsuhashi, 1981; Schilperoord, 2002). This is particularly reflected in the model proposed by Flower and Hayes (1980) that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts (also Chenoweth \& Hayes, 2001, 2003).

This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses (Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002). However this account is not consistent with what we know about writing. Consider these two observations:

First, utterances are not fully formed at production onset. Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process. This is well known for spoken production (e.g. Bock \& Ferreira, 2014). For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases {[}Nottbusch (2010); dam09; Roeser et al. (2019){]}. For example, in Roeser et al. (2019) participants described moving arrays of images in simple utterances such as \emph{A and the B moved above the C} that either started with a conjoined noun phrase (i.e.~\emph{A and the B moved}) or a simple noun (i.e., \emph{A moved above the B and the C}). Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset. This is not what one would expect if writers plan sentences (or even phrases) in full before production onset. In fact, eye movement data reported in Torrance and Nottbusch (2012) and Roeser et al. (2019) demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset (similar to speech Griffin, 2001, 2003), although there was evidence that lexical pre-planning of the second noun (i.e.~\emph{the B}) was more likely for conjoined noun phrases.

Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid. For example, Medimorec and Risko (2017) found that, in undergraduate students writing on familiar topics 71\% of sentence-initial IKIs were less than 1 sec in duration. Also Rønneberg et al. (2022) reported that sentence-initial hesitations are rare, with over 50\% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses. For comparison, this is less than mean written naming response latency for single objects (Torrance et al., 2018) and short sentences when describing arrays of images (Roeser et al., 2019) in a similar population. Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent. The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.

If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations. The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control. Consistent with general trends in language processing theory (Bock \& Ferreira, 2014; Chang et al., 2006; Dell et al., 2007), several researchers have argued that the processes associated with written production run at least in part in parallel (Bonin et al., 2012; Crump \& Logan, 2010; Olive, 2014; Roux et al., 2013; Van Galen, 1991). Van Galen (1991), for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available (Christiansen \& Chater, 2016). Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes. In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.

This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals. IKIs result from one of two data-generating processes. When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e.~just by processing below the last buffer in the processing cascade). However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement. IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels (Roeser et al., 2021).

For example, Roeser et al. (2021) demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling. In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach. Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions. As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition. Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes. For similar approaches see Almond et al. (2012), Baaijen et al. (2012), Chenu et al. (2014), Guo et al. (2018), Hall et al. (2022), Li (2021), and Van Waes et al. (2021).

The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not. This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.

Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. Rønneberg et al. (2022) coined this possibility the \enquote{process-disruption hypothesis.}

In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to Roeser et al. (2021) -- whether the cascading view of writing generalises to contexts in which participants compose texts. On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature (Ailhaud et al., 2016; Ailhaud \& Chenu, 2017; Chukharev-Hudilainen et al., 2019; Conijn et al., 2019; Mohsen, 2021; Torrance, Rønneberg, et al., 2016; Wengelin, 2002).

Data form spontaneous text production (see Gernsbacher \& Givón, 1995) are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production. Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy (Dux Speltz et al., 2022; Dux Speltz \& Chukharev-Hudilainen, 2021; Vandermeulen et al., 2023).

Second, since the emergence of software such as InputLog (Leijten \& Van Waes, 2013) recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing (Conijn \& Torrance, 2023). Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016. These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning (Conijn et al., 2019; Conijn, 2020).

Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments (Damian \& Stadthagen-Gonzalez, 2009; Nottbusch, 2010; Roeser et al., 2019) or copy tasks (Roeser et al., 2021; Van Waes et al., 2019, 2021). In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks. Instead text composition provides data from both inter and intra-sentence transitions. Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).

The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location. In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations. This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation. The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing. In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production. Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability. We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models. The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models. In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).

The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in Roeser et al. (2021). Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow. Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur. This view was implemented as a constrained and an unconstrained model.

\hypertarget{statistical-models}{%
\subsection{Statistical models}\label{statistical-models}}

We are using the Bayesian framework (e.g. Farrell \& Lewandowsky, 2018; Gelman et al., 2014; M. D. Lee \& Wagenmakers, 2014) to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing. In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models. The models presented in this section build on one another so that later model include assumptions of earlier models.

\hypertarget{serial-model-of-writing}{%
\subsubsection{Serial model of writing}\label{serial-model-of-writing}}

\hypertarget{unimodal-gaussian}{%
\paragraph{Unimodal Gaussian}\label{unimodal-gaussian}}

Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit. The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors. For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes. There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis. We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean \(\mu\), that describes the average IKI associated with word-level planning, and a standard deviation \(\sigma_\text{e}^2\), that captures the variability associated with factors that we did not further specify in the model. This can be expressed as \(\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)\). Of importance is the estimated posterior value of \(\mu\) as this value captures that time it takes to mentally plan a word.

We can extend this simple model of word-planning to other linguistic location. We introduce earlier that larger linguistic edges are associated with planning on higher levels. For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI \(\mu\) -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun (Nottbusch et al., 2007; Roeser et al., 2019). Regression formulas can, and commonly are, used to decompose \(\mu\) and capture that there is a change in the outcome variable associated with another factors. We can decompose \(\mu\) as \(\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}\) so that when \(\text{x}_\text{sentence}\) takes on the value 0, the equation reduces to \(\mu = \alpha\) which is the average IKI for word boundaries. However when \(\text{x}_\text{sentence}\) takes on the value 1, the average IKI for word boundaries \(\alpha\) is incremented by a sentence-related slowdown of \(\beta\) msecs. Therefore the value of the \(\beta\) parameter represents the additional cognitive demand associated with sentence-initial planning. The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g.~whether there is a statistically meaningful difference between IKI associated with words and sentences).

For computational ease, we implemented the differences associated with transition locations as \(\beta_\text{location}\) in equation \ref{eq:unimodgaus} where \(\text{location}\) is taking on an index for each transition location. Therefore, the model will return one \(\beta\) per transition location that capture the posterior distribution of average IKIs. Also the decomposition of \(\mu\) allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts \(u_\text{participant}\). The random intercepts term \(u_\text{participant}\) is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of \(\sigma_\text{p}^2\). This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e.~a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).

The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.

\begin{equation}
\begin{aligned}
\label{eq:unimodgaus}
\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.

\hypertarget{unimodal-log-gaussian}{%
\paragraph{Unimodal log-Gaussian}\label{unimodal-log-gaussian}}

The previous model assumes that the data-generating process is a Gaussian distribution. The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution. There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values. IKIs, as the distance between two subsequent key-down events, must be positive. The lower bound is delimited by a person's ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human behaviour and motor responses. In particular, a Gaussian distribution assumes that units are scaled linearly. For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs. This does not necessarily map onto the psychological interpretation for short and long keyintervals. For example effects that result on the motor level within words (e.g.~typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2). In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals. Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).

The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution \(\text{N}()\) was replaced by \(\text{logN}()\).

\begin{equation}
\begin{aligned}
\label{eq:unimodloggaus}
\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

\hypertarget{unimodal-unequal-variance-log-gaussian}{%
\paragraph{Unimodal unequal-variance log-Gaussian}\label{unimodal-unequal-variance-log-gaussian}}

The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section. The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text. The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance). This assumption is however not in line with what we know about data from human behaviour. Longer latencies are known to be associated with a larger variances for both response-time data in particular (Wagenmakers \& Brown, 2007) and human motor behaviour in general (Schöner, 2002; Wing \& Kristofferson, 1973). For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes. Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation \(\sigma_{e_\text{location}}^2)\) varies as a function of transition location.

\begin{equation}
\begin{aligned}
\label{eq:unimoduv}
\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
 & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

\hypertarget{parallel-cascading-model-of-writing}{%
\subsubsection{Parallel cascading model of writing}\label{parallel-cascading-model-of-writing}}

The following two models are are extensions of the models introduced for the serial view. Crucially, the cascading view allows planning to happen in parallel to production. Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset. This is done by assuming that IKIs come from a weighted combination of two distributions.

\hypertarget{bimodal-log-gaussian-constrained}{%
\paragraph{Bimodal log-Gaussian (constrained)}\label{bimodal-log-gaussian-constrained}}

This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses. Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory. This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Activation can flow into keystrokes without interruption. These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the \(\beta\) parameter in equation \ref{eq:bimodcon}. Note that \(\beta\) in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.
\item
  Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time. The slowdown for these hesitant transitions is captured by \(\delta\) in the first line of equation \ref{eq:bimodcon}. The magnitude of the slowdown is associated with transition location. This is because delays at larger linguistic units are likely to be associated with higher level planning. By constraining \(\delta\) to be positive, it captures how much longer hesitant IKIs in addition to \(\beta\).
\end{enumerate}

\begin{equation}
\begin{aligned}
\label{eq:bimodcon}
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
        & \sigma_{e'}^2 > \sigma_{e}^2\\
        & 0 > \theta > 1
\end{aligned}
\end{equation}

The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions. Each of these two distributions is associated with the mixing weight \(\theta\) which is a proportion that is constrained to be larger than 0 and smaller than 1. \(\theta\) is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs. This probability is inversely related to the mixing weight of the distribution of short IKIs by \(1-\theta\). In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution. The weights of both distributions must sum to 1. We will call this parameter the probability of hesitant transitions.

The probability of hesitant transitions is assumed to vary as a function of both transition location and participants. In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges. As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less. This is akin to what is known as a random by-participant slopes model.

>>>>

\begin{appendix}
\section{}
\hypertarget{transition-location-effect-unconstrained-mixture-model}{%
\subsection{Transition location effect (unconstrained mixture
model)}\label{transition-location-effect-unconstrained-mixture-model}}

We present the mixture-model posterior estimates for each of the three
parameter values in the three facets of Figure \ref{fig:crossstudypost}.
Similar to the constrained mixture model, we aggregated the posterior
across conditions\footnote{We aggregated across pre-post test for the
  PLanTra dataset as well as genre and topic of the LIFT data set.} and
removed conditions that might conflate comparisons\footnote{We removed
  the masked writing condition in the GUNNEXP2 and CATO, the dyslexic
  group in the CATO data set, and L2 writing in the SPL2 data set.}. The
resulting posterior allows us to examine differences between transition
locations for each data set associated with each of the three
mixture-model parameters.
